Different rules for binocular combination of luminance flicker in cortical and subcortical pathways

### 2023-02-08 manuscriptPLOSBIO.tex
Edited for PLOS Biology

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Different rules for binocular combination of luminance flicker in cortical and subcortical pathways},
  pdfauthor={Federico G. Segala, Aurelio Bruno, Myat T. Aung, Alex R. Wade \& Daniel H. Baker},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}\doublespacing

@@

Federico G. Segala, email: \href{mailto:fgs502@york.ac.uk}{\nolinkurl{fgs502@york.ac.uk}}

\textbf{Author contributions: Federico Segala}: Methodology, software, formal analysis, investigation, data curation, writing - original draft, writing - review \& editing, visualization. \textbf{Aurelio Bruno}: Conceptualization, writing - review \& editing, supervision, project administration, funding acquisition. \textbf{Myat Aung}: Software, resources, writing - review \& editing. \textbf{Alex Wade}: Conceptualization, methodology, resources, writing - review \& editing, supervision, project administration, funding acquisition. \textbf{Daniel Baker}: Conceptualization, methodology, software, formal analysis, investigation, resources, data curation, writing - original draft, writing - review \& editing, visualization, supervision, project administration, funding acquisition.

\textbf{Competing interest declaration:} The authors declare no competing interests.

\textbf{Classification:} Biological Sciences - Neuroscience

\textbf{Keywords:} Pupillary light reflex; Pupillometry; EEG; Pupil diameter; Binocular vision; Binocular interactions; Contrast matching; Computational modelling; Psychophysics

@@

\hypertarget{significance-statement}{%
\section{Significance Statement}\label{significance-statement}}

Please note that it should not exceed 120 words, but should be at least 50 words in length.

@@

@@

All stimuli had a mean luminance of 42 cd/m\(^2\), and were presented on an Iiyama VisionMaster™ Pro 510 display (800 x 600 pixels, 60 Hz refresh rate), which was gamma corrected using a Minolta LS-110 photometer (Minolta Camera Co.~Ltd., Japan). All stimuli were presented using Psychopy {[}20{]} (v3.0.7 for Experiments 1 and 2; v2022.1.1 for Experiment 3).

EEG data were collected for Experiments 1 and 2 using a 64 electrode ANT WaveGuard cap and the signals were recorded at 1 kHz using the ASA software (ANT Neuro, Netherlands). Pupillometry data were collected for Experiment 1 using a binocular Pupil Core eye-tracker (Pupil Labs GmbH, Berlin, Germany {[}21{]}) running at 120 Hz, and the signals were recorded with the Pupil Capture software.

@@

EEG data were converted from the ANT-EEProbe format to a compressed csv text file using a custom Matlab script and components of the EEGlab toolbox {[}22{]}. The data for each participant were then loaded into R for analysis, where a ten-second waveform for each trial at each electrode was extracted (omitting the first two seconds). The Fourier transform of each waveform was calculated, and the complex spectrum stored in a matrix. All repetitions of each condition were then averaged for each electrode. They were then averaged across four occipital electrodes (\emph{POz}, \emph{Oz}, \emph{O1}, \emph{O2}), to obtain individual results. Finally, these were averaged across participants to obtain the group results. All averaging was performed in the complex domain and therefore retained the phase information (i.e.~coherent averaging), and at each stage we excluded data points with a Mahalanobis distance exceeding \(D\) = 3 from the complex-valued mean {[}23{]}. For statistical comparisons of complex-valued data, we use the \(ANOVA^2_{circ}\) statistic described previously {[}23{]}. This is a multivariate extension of ANOVA that assumes equal variance of the real and imaginary Fourier components, or equivalently, an extension of the \(T^2_{circ}\) statistic {[}24{]} that can compare more than two conditions.

A similar analysis pipeline was adopted for the pupillometry data. The data were converted from mp4 videos to a csv text file using the Pupil Player software {[}21{]}, which estimated pupil diameter for each eye on each frame using a 3D model of the eyeball. The individual data were then loaded into R for analysis, where again a ten-second waveform for each trial in each eye was extracted (excluding the first two seconds after stimulus onset). We interpolated across any dropped or missing frames to ensure regular and continuous sampling over time. The Fourier transform was calculated for each waveform, and all repetitions of each condition were pooled across eye and then averaged. Finally, data were averaged across all participants to obtain the group results. Again, we used coherent averaging, and excluded outlying data points in the same way as for the EEG data. Note that previous pupillometry studies using luminance flicker have tended to fit a single sine-wave at the fundamental frequency, rather than using Fourier analysis {[}19{]}. The Fourier approach is more robust to noise at other frequencies and has been used in some previous studies {[}25,26{]}. Additionally, it makes the pupillometry analysis consistent with standard practice in steady-state EEG analysis {[}27{]}.

@@

Despite being derived from the model proposed by Meese et al.~{[}9{]}, the simplifications applied to this architecture make it very similar to other models {[}8,11,17,28{]}. In particular we fixed the numerator exponent at 2 in our model, because otherwise this value tends to trade off with the weight of interocular suppression {[}29,30{]}. Our key parameter of interest is the weight of interocular suppression. Large values around \(w\) = 1 result in a very small or nonexistent binocular advantage at suprathreshold contrasts, consistent with previous work using grating stimuli {[}1{]}. Low values around \(w\) = 0 produce substantial, near-linear binocular facilitation {[}31{]}.

We implemented the model within a Bayesian framework using the Stan software {[}32{]}. This allowed us to estimate group-level posterior parameter distributions for the weight of interocular suppression, \(w\), and the other free model parameters \(R_{max}\), \(Z\) and \(n\). The prior distributions for all parameters were Gaussian, with means and standard deviations of 1 and 0.5 for \(w\) and \(R_{max}\), and 5 and 2 for \(Z\) and \(n\), with these values chosen based on previous literature {[}9,29{]}. We sampled from a Student's t-distribution for the amplitudes in the pupillometry and EEG experiments, and from a Bernoulli distribution for the single trial matching data. The models were fit using the individual data across all participants, independently for each data set. Because we used coherent averaging across participants for our steady-state data, the group average amplitudes are shifted vertically relative to the model predictions, which are based on hierarchical fits to the individual participant amplitudes (put another way, the model does not implement coherent averaging). However it is clear that the character of the model in all cases gives a good representation of the data, as can be seen in Fig. \ref{fig:modelfigure}. We took posterior samples at over a million steps for each data set, using a computer cluster, and retained 10\% of samples for plotting.

@@

{\centering \includegraphics{Figures/pupildata}

Fig. \ref{fig:pupildata}c shows contrast response functions in response to stimuli flickering only at 2Hz. Response amplitudes increased monotonically with target contrast, confirming that our paradigm is suitable for measuring contrast-dependent differences in response (to our knowledge this is the first time this has been demonstrated). The amplitude of the binocular condition (blue squares) is consistently greater than that of the monocular condition (red circles) across all target contrasts. A \(2\times5\) repeated measures \(ANOVA^2_{circ}\) {[}23{]} comparing these conditions revealed a significant main effect of target contrast (F(8,580) = 16.79, \(p\) \textless{} 0.001), a significant effect of condition (F(2,580) = 11.04, \(p\) \textless{} 0.001), and a significant interaction (F(8,580) = 56.25, \(p\) \textless{} 0.001). The dichoptic condition begins at a much higher amplitude, owing to binocular combination of the target and high (48\%) contrast mask, and then increases slightly with increasing target contrast (main effect of target contrast: F(8,232) = 3.03, \(p\) \textless{} 0.003).

{\centering \includegraphics{Figures/EEGdata}

{\centering \includegraphics[width=0.5\linewidth]{Figures/BSratios}

{\centering \includegraphics{Figures/TFdata}

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms {[}33--35{]}. We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

{\centering \includegraphics[width=0.5\linewidth]{Figures/matchingdata}

{\centering \includegraphics{Figures/modelfigure}

The results of our main experiment were unexpected for both the pupillometry and the EEG measures. Previous studies investigating binocular combination of spatial patterns (i.e.~sine wave grating stimuli) are universally consistent with strong interocular suppression and weak binocular facilitation at high contrasts {[}1,9,10{]}. Our second experiment ruled out the possibility that these differences were due to the lower temporal frequency (2Hz) used here. However, there is evidence of more extensive binocular facilitation for a range of other stimuli. For example, Quaia et al.@Quaia2018 observed a strong binocular facilitation (or `supersummation') in the reflexive eye movement response to rapidly moving stimuli, and Spitschan and \href{mailto:Cajochen@Spitschan2019}{\nolinkurl{Cajochen@Spitschan2019}} report a similar result in archival data on melatonin suppression due to light exposure. In the auditory system, interaural suppression of amplitude modulation also appears to be weak when measured using a similar steady-state paradigm {[}31{]}. Finally, psychophysical matching experiments using static stimuli also show near-linear behaviour for luminance increments {[}29,33,34{]}, though not for luminance decrements {[}33{]}. Overall, this suggests that strong interocular normalization may be specific to spatial pattern vision, and not a general feature of binocular signal combination.

Given the above, where does this leave our understanding of the overarching purpose of signal combination? We \href{mailto:previously@Baker2017}{\nolinkurl{previously@Baker2017}} pointed out that a model with strong suppression between channels that are subsequently summed is equivalent to a Kalman filter, which is the optimal method for combining two noisy inputs (see also {[}2{]}). This account has intuitive appeal, and is consistent with other models that propose binocular combination as a means of redundancy reduction {[}36,37{]}. One possibility is that optimal combination is useful for visual perception --- a critical system for interacting with the local environment --- and is therefore worth devoting the additional resource of inhibitory wiring between ocular channels. However the other examples given in the previous paragraph are primarily physiological responses (pupil size, eye movements, hormone release) that may benefit less from an increased signal-to-noise ratio, or are otherwise phylogenetically older than binocular pattern vision. Conceptualised another way, the brain can repurpose a generic architecture for different situational demands by adjusting parameter values (here the weight of interocular suppression) to achieve different outcomes. Our future work in this area intends to compare binocular combination for specific photoreceptor pathways, including different cone classes, and intrinsically photoreceptive retinal ganglion cells.

Pupil size determines the total amount of light falling on the retina. It is therefore the case that fluctuations in pupil diameter have a downstream effect on the signals reaching cortex. We did not incorporate such interactions into our computational model, though in principle this might be worthwhile. However we anticipate than any such effects would be small, since pupil modulations at 2Hz are in the order of 2\% of overall diameter {[}19{]}. It is also the case that cortical activity can modulate pupil diameter, usually through arousal and attention mechanisms {[}38{]}. We think it unlikely that these temporally coarse processes would have a differential effect on e.g.~monocular and binocular stimulation conditions in our experiment, and any fluctuations during an experimental session (perhaps owing to fatigue) will be equivalent for our comparisons of interest. Therefore we make the simplifying assumption that the two pathways are effectively distinct, but intend to investigate this directly in future neuroimaging work.

Classic studies investigating the neurophysiological architecture of V1 reported that cells in cytochrome-oxidase `blobs' {[}39,40{]} are biased towards low spatial frequencies {[}41,42{]}, and relatively insensitive to stimulus orientation {[}39,40{]} (though see {[}43{]}). As the blob regions are embedded within ocular dominance columns {[}39{]}, they are also largely monocular {[}40,44{]}. More recent work has reported psychophysical evidence for unoriented chromatic {[}45{]} and achromatic {[}46{]} mechanisms, that also appear to be monocular. Our use of luminance flicker might preferentially stimulate these mechanisms, perhaps explaining why our EEG data show little evidence of binocular interactions. Indeed, our EEG results could potentially be explained by a model involving entirely non-interacting monocular channels, with the binocular facilitation effects we find (e.g.~Fig. \ref{fig:BSratios} \& \ref{fig:TFdata}) owing to additivity of the electrophysiological response across independent monocular cells, rather than requiring binocular neurons. However our matching data (Fig. \ref{fig:matchingdata}), as well as everyday perceptual experience, indicates that luminance signals must be combined physiologically at some stage, and that this process still involves weak or absent interocular suppression.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

Supported by Biotechnology and Biological Sciences Research Council grant BB/V007580/1 awarded to DHB and ARW, and Wellcome Trust grant 213616/Z/18/Z to AB.

@@

\leavevmode\hypertarget{ref-Peirce2019}{}%
20. Peirce J, Gray JR, Simpson S, MacAskill M, Höchenberger R, Sogo H, et al. PsychoPy2: Experiments in behavior made easy. Behav Res Methods. 2019;51: 195--203. doi:\href{https://doi.org/10.3758/s13428-018-01193-y}{10.3758/s13428-018-01193-y}

\leavevmode\hypertarget{ref-Kassner2014}{}%
21. Kassner M, Patera W, Bulling A. Pupil: An open source platform for pervasive eye tracking and mobile gaze-based interaction. Proceedings of the 2014 ACM international joint conference on pervasive and ubiquitous computing: Adjunct publication. UbiComp '14 Adjunct; ACM; 2014. pp. 1151--1160. doi:\href{https://doi.org/10.1145/2638728.2641695}{10.1145/2638728.2641695}

\leavevmode\hypertarget{ref-Delorme2004}{}%
22. Delorme A, Makeig S. EEGLAB: An open source toolbox for analysis of single-trial EEG dynamics including independent component analysis. J Neurosci Methods. 2004;134: 9--21. doi:\href{https://doi.org/10.1016/j.jneumeth.2003.10.009}{10.1016/j.jneumeth.2003.10.009}

\leavevmode\hypertarget{ref-Baker2021}{}%
23. Baker DH. Statistical analysis of periodic data in neuroscience. Neurons, Behavior, Data analysis, and Theory. 2021;5. doi:\href{https://doi.org/10.51628/001c.27680}{10.51628/001c.27680}

\leavevmode\hypertarget{ref-Victor1991}{}%
24. Victor JD, Mast J. A new statistic for steady-state evoked potentials. Electroencephalogr Clin Neurophysiol. 1991;78: 378--88. doi:\href{https://doi.org/10.1016/0013-4694(91)90099-p}{10.1016/0013-4694(91)90099-p}

\leavevmode\hypertarget{ref-Barrionuevo2014}{}%
25. Barrionuevo PA, Nicandro N, McAnany JJ, Zele AJ, Gamlin P, Cao D. Assessing rod, cone, and melanopsin contributions to human pupil flicker responses. Invest Ophthalmol Vis Sci. 2014;55: 719--27. doi:\href{https://doi.org/10.1167/iovs.13-13252}{10.1167/iovs.13-13252}

\leavevmode\hypertarget{ref-Barrionuevo2016}{}%
26. Barrionuevo PA, Cao D. Luminance and chromatic signals interact differently with melanopsin activation to control the pupil light response. J Vis. 2016;16(11): 29. doi:\href{https://doi.org/10.1167/16.11.29}{10.1167/16.11.29}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Different rules for binocular combination of luminance flicker in cortical and subcortical pathways},
  pdfauthor={Federico G. Segala, Aurelio Bruno, Myat T. Aung, Alex R. Wade \& Daniel H. Baker},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{setspace}\doublespacing
\usepackage{lineno}\linenumbers

@@

\textbf{Short title}: ``Binocular combination of flickering light''

\textbf{Corresponding author}: Federico G. Segala, email: \href{mailto:fgs502@york.ac.uk}{\nolinkurl{fgs502@york.ac.uk}}
\newpage

@@

@@

{\centering \includegraphics{Figures/blank}

Fig. \ref{fig:pupildata}c shows contrast response functions in response to stimuli flickering only at 2Hz. Response amplitudes increased monotonically with target contrast, confirming that our paradigm is suitable for measuring contrast-dependent differences in response (to our knowledge this is the first time this has been demonstrated). The amplitude of the binocular condition (blue squares) is consistently greater than that of the monocular condition (red circles) across all target contrasts. A \(2\times5\) repeated measures \(ANOVA^2_{circ}\) {[}20{]} comparing these conditions revealed a significant main effect of target contrast (F(8,580) = 16.79, \(p\) \textless{} 0.001), a significant effect of condition (F(2,580) = 11.04, \(p\) \textless{} 0.001), and a significant interaction (F(8,580) = 56.25, \(p\) \textless{} 0.001). The dichoptic condition begins at a much higher amplitude, owing to binocular combination of the target and high (48\%) contrast mask, and then increases slightly with increasing target contrast (main effect of target contrast: F(8,232) = 3.03, \(p\) \textless{} 0.003).

{\centering \includegraphics{Figures/blank}

{\centering \includegraphics[width=0.5\linewidth]{Figures/blank}

{\centering \includegraphics{Figures/blank}

In Experiment 1 we found evidence of stronger binocular facilitation for cortical responses to luminance flicker (measured using EEG), compared with subcortical responses (measured using pupillometry). Since perception is dependent on cortical responses, these results provide a clear prediction for perceived contrast judgements indexed by psychophysical contrast matching paradigms {[}21--23{]}. We therefore conducted such an experiment, in which participants judged which of two stimuli had the greater perceived amplitude of flicker. One stimulus was a matching stimulus, that had a fixed binocular flicker amplitude of either 24\% or 48\% (temporal) contrast. The other stimulus was a target stimulus, the contrast of which was controlled by a staircase algorithm. We tested 9 ratios of contrast between the left and right eyes.

{\centering \includegraphics[width=0.5\linewidth]{Figures/blank}

{\centering \includegraphics{Figures/blank}

The results of our main experiment were unexpected for both the pupillometry and the EEG measures. Previous studies investigating binocular combination of spatial patterns (i.e.~sine wave grating stimuli) are universally consistent with strong interocular suppression and weak binocular facilitation at high contrasts {[}1,9,10{]}. Our second experiment ruled out the possibility that these differences were due to the lower temporal frequency (2Hz) used here. However, there is evidence of more extensive binocular facilitation for a range of other stimuli. For example, Quaia et al.@Quaia2018 observed a strong binocular facilitation (or `supersummation') in the reflexive eye movement response to rapidly moving stimuli, and Spitschan and \href{mailto:Cajochen@Spitschan2019}{\nolinkurl{Cajochen@Spitschan2019}} report a similar result in archival data on melatonin suppression due to light exposure. In the auditory system, interaural suppression of amplitude modulation also appears to be weak when measured using a similar steady-state paradigm {[}24{]}. Finally, psychophysical matching experiments using static stimuli also show near-linear behaviour for luminance increments {[}21,22,25{]}, though not for luminance decrements {[}21{]}. Overall, this suggests that strong interocular normalization may be specific to spatial pattern vision, and not a general feature of binocular signal combination.

Given the above, where does this leave our understanding of the overarching purpose of signal combination? We \href{mailto:previously@Baker2017}{\nolinkurl{previously@Baker2017}} pointed out that a model with strong suppression between channels that are subsequently summed is equivalent to a Kalman filter, which is the optimal method for combining two noisy inputs (see also {[}2{]}). This account has intuitive appeal, and is consistent with other models that propose binocular combination as a means of redundancy reduction {[}26,27{]}. One possibility is that optimal combination is useful for visual perception --- a critical system for interacting with the local environment --- and is therefore worth devoting the additional resource of inhibitory wiring between ocular channels. However the other examples given in the previous paragraph are primarily physiological responses (pupil size, eye movements, hormone release) that may benefit less from an increased signal-to-noise ratio, or are otherwise phylogenetically older than binocular pattern vision. Conceptualised another way, the brain can repurpose a generic architecture for different situational demands by adjusting parameter values (here the weight of interocular suppression) to achieve different outcomes. Our future work in this area intends to compare binocular combination for specific photoreceptor pathways, including different cone classes, and intrinsically photoreceptive retinal ganglion cells.

Pupil size determines the total amount of light falling on the retina. It is therefore the case that fluctuations in pupil diameter have a downstream effect on the signals reaching cortex. We did not incorporate such interactions into our computational model, though in principle this might be worthwhile. However we anticipate than any such effects would be small, since pupil modulations at 2Hz are in the order of 2\% of overall diameter {[}19{]}. It is also the case that cortical activity can modulate pupil diameter, usually through arousal and attention mechanisms {[}28{]}. We think it unlikely that these temporally coarse processes would have a differential effect on e.g.~monocular and binocular stimulation conditions in our experiment, and any fluctuations during an experimental session (perhaps owing to fatigue) will be equivalent for our comparisons of interest. Therefore we make the simplifying assumption that the two pathways are effectively distinct, but intend to investigate this directly in future neuroimaging work.

Classic studies investigating the neurophysiological architecture of V1 reported that cells in cytochrome-oxidase `blobs' {[}29,30{]} are biased towards low spatial frequencies {[}31,32{]}, and relatively insensitive to stimulus orientation {[}29,30{]} (though see {[}33{]}). As the blob regions are embedded within ocular dominance columns {[}29{]}, they are also largely monocular {[}30,34{]}. More recent work has reported psychophysical evidence for unoriented chromatic {[}35{]} and achromatic {[}36{]} mechanisms, that also appear to be monocular. Our use of luminance flicker might preferentially stimulate these mechanisms, perhaps explaining why our EEG data show little evidence of binocular interactions. Indeed, our EEG results could potentially be explained by a model involving entirely non-interacting monocular channels, with the binocular facilitation effects we find (e.g.~Fig. \ref{fig:BSratios} \& \ref{fig:TFdata}) owing to additivity of the electrophysiological response across independent monocular cells, rather than requiring binocular neurons. However our matching data (Fig. \ref{fig:matchingdata}), as well as everyday perceptual experience, indicates that luminance signals must be combined physiologically at some stage, and that this process still involves weak or absent interocular suppression.

@@

All stimuli had a mean luminance of 42 cd/m\(^2\), and were presented on an Iiyama VisionMaster™ Pro 510 display (800 x 600 pixels, 60 Hz refresh rate), which was gamma corrected using a Minolta LS-110 photometer (Minolta Camera Co.~Ltd., Japan). All stimuli were presented using Psychopy {[}37{]} (v3.0.7 for Experiments 1 and 2; v2022.1.1 for Experiment 3).

EEG data were collected for Experiments 1 and 2 using a 64 electrode ANT WaveGuard cap and the signals were recorded at 1 kHz using the ASA software (ANT Neuro, Netherlands). Pupillometry data were collected for Experiment 1 using a binocular Pupil Core eye-tracker (Pupil Labs GmbH, Berlin, Germany {[}38{]}) running at 120 Hz, and the signals were recorded with the Pupil Capture software.

@@

EEG data were converted from the ANT-EEProbe format to a compressed csv text file using a custom Matlab script and components of the EEGlab toolbox {[}39{]}. The data for each participant were then loaded into R for analysis, where a ten-second waveform for each trial at each electrode was extracted (omitting the first two seconds). The Fourier transform of each waveform was calculated, and the complex spectrum stored in a matrix. All repetitions of each condition were then averaged for each electrode. They were then averaged across four occipital electrodes (\emph{POz}, \emph{Oz}, \emph{O1}, \emph{O2}), to obtain individual results. Finally, these were averaged across participants to obtain the group results. All averaging was performed in the complex domain and therefore retained the phase information (i.e.~coherent averaging), and at each stage we excluded data points with a Mahalanobis distance exceeding \(D\) = 3 from the complex-valued mean {[}20{]}. For statistical comparisons of complex-valued data, we use the \(ANOVA^2_{circ}\) statistic described previously {[}20{]}. This is a multivariate extension of ANOVA that assumes equal variance of the real and imaginary Fourier components, or equivalently, an extension of the \(T^2_{circ}\) statistic {[}40{]} that can compare more than two conditions.

A similar analysis pipeline was adopted for the pupillometry data. The data were converted from mp4 videos to a csv text file using the Pupil Player software {[}38{]}, which estimated pupil diameter for each eye on each frame using a 3D model of the eyeball. The individual data were then loaded into R for analysis, where again a ten-second waveform for each trial in each eye was extracted (excluding the first two seconds after stimulus onset). We interpolated across any dropped or missing frames to ensure regular and continuous sampling over time. The Fourier transform was calculated for each waveform, and all repetitions of each condition were pooled across eye and then averaged. Finally, data were averaged across all participants to obtain the group results. Again, we used coherent averaging, and excluded outlying data points in the same way as for the EEG data. Note that previous pupillometry studies using luminance flicker have tended to fit a single sine-wave at the fundamental frequency, rather than using Fourier analysis {[}19{]}. The Fourier approach is more robust to noise at other frequencies and has been used in some previous studies {[}41,42{]}. Additionally, it makes the pupillometry analysis consistent with standard practice in steady-state EEG analysis {[}43{]}.

@@

Despite being derived from the model proposed by Meese et al.~{[}9{]}, the simplifications applied to this architecture make it very similar to other models {[}8,11,17,44{]}. In particular we fixed the numerator exponent at 2 in our model, because otherwise this value tends to trade off with the weight of interocular suppression {[}25,45{]}. Our key parameter of interest is the weight of interocular suppression. Large values around \(w\) = 1 result in a very small or nonexistent binocular advantage at suprathreshold contrasts, consistent with previous work using grating stimuli {[}1{]}. Low values around \(w\) = 0 produce substantial, near-linear binocular facilitation {[}24{]}.

We implemented the model within a Bayesian framework using the Stan software {[}46{]}. This allowed us to estimate group-level posterior parameter distributions for the weight of interocular suppression, \(w\), and the other free model parameters \(R_{max}\), \(Z\) and \(n\). The prior distributions for all parameters were Gaussian, with means and standard deviations of 1 and 0.5 for \(w\) and \(R_{max}\), and 5 and 2 for \(Z\) and \(n\), with these values chosen based on previous literature {[}9,25{]}. We sampled from a Student's t-distribution for the amplitudes in the pupillometry and EEG experiments, and from a Bernoulli distribution for the single trial matching data. The models were fit using the individual data across all participants, independently for each data set. Because we used coherent averaging across participants for our steady-state data, the group average amplitudes are shifted vertically relative to the model predictions, which are based on hierarchical fits to the individual participant amplitudes (put another way, the model does not implement coherent averaging). However it is clear that the character of the model in all cases gives a good representation of the data, as can be seen in Fig. \ref{fig:modelfigure}. We took posterior samples at over a million steps for each data set, using a computer cluster, and retained 10\% of samples for plotting.

@@

\hypertarget{author-contributions}{%
\section{Author contributions}\label{author-contributions}}

\textbf{Federico Segala}: Methodology, software, formal analysis, investigation, data curation, writing - original draft, writing - review \& editing, visualization. \textbf{Aurelio Bruno}: Conceptualization, writing - review \& editing, supervision, project administration, funding acquisition. \textbf{Myat Aung}: Software, resources, writing - review \& editing. \textbf{Alex Wade}: Conceptualization, methodology, resources, writing - review \& editing, supervision, project administration, funding acquisition. \textbf{Daniel Baker}: Conceptualization, methodology, software, formal analysis, investigation, resources, data curation, writing - original draft, writing - review \& editing, visualization, supervision, project administration, funding acquisition.

@@

\leavevmode\hypertarget{ref-Baker2021}{}%
20. Baker DH. Statistical analysis of periodic data in neuroscience. Neurons, Behavior, Data analysis, and Theory. 2021;5. doi:\href{https://doi.org/10.51628/001c.27680}{10.51628/001c.27680}

\leavevmode\hypertarget{ref-Anstis1998}{}%
21. Anstis S, Ho A. Nonlinear combination of luminance excursions during flicker, simultaneous contrast, afterimages and binocular fusion. Vision Res. 1998;38: 523--39. doi:\href{https://doi.org/10.1016/s0042-6989(97)00167-3}{10.1016/s0042-6989(97)00167-3}

\leavevmode\hypertarget{ref-Levelt1965}{}%
22. Levelt WJ. Binocular brightness averaging and contour information. Br J Psychol. 1965;56: 1--13. doi:\href{https://doi.org/10.1111/j.2044-8295.1965.tb00939.x}{10.1111/j.2044-8295.1965.tb00939.x}

\leavevmode\hypertarget{ref-Quaia2018}{}%
23. Quaia C, Optican LM, Cumming BG. Binocular summation for reflexive eye movements. J Vis. 2018;18(4): 7. doi:\href{https://doi.org/10.1167/18.4.7}{10.1167/18.4.7}

\leavevmode\hypertarget{ref-Baker2020}{}%
24. Baker DH, Vilidaite G, McClarnon E, Valkova E, Bruno A, Millman RE. Binaural summation of amplitude modulation involves weak interaural suppression. Sci Rep. 2020;10: 3560. doi:\href{https://doi.org/10.1038/s41598-020-60602-5}{10.1038/s41598-020-60602-5}

\leavevmode\hypertarget{ref-Baker2012}{}%
25. Baker DH, Wallis SA, Georgeson MA, Meese TS. Nonlinearities in the binocular combination of luminance and contrast. Vision Res. 2012;56: 1--9. doi:\href{https://doi.org/10.1016/j.visres.2012.01.008}{10.1016/j.visres.2012.01.008}

\leavevmode\hypertarget{ref-Li1994}{}%
26. Li Z, Atick JJ. Efficient stereo coding in the multiscale representation. Network: Computation in Neural Systems. 1994;5: 157--174. doi:\href{https://doi.org/10.1088/0954-898X_5_2_003}{10.1088/0954-898X\_5\_2\_003}

\leavevmode\hypertarget{ref-May2022}{}%
27. May K, Zhaoping L. Li and Atick's theory of efficient binocular coding: A tutorial and mini-review. Vision Res. 2022;201: 107950. doi:\href{https://doi.org/10.1016/j.visres.2021.08.005}{10.1016/j.visres.2021.08.005}
